CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-04-07 script/en/chapters/primes.tex
more fixes
richard's comments

This number was discovered as 39th Mersenne prime (and already called M-39,
despite it has not been proven yet, whether no further Mersenne prime
numbers between M-38 und M13466917 do indeed exist), $$2^{13,466,917}-1,$$
at December 6, 2001 -- more exactly, the verification of this number,
found at November 14, 2001 by the Canadian student Michael Cameron, was
successfully completed. This number has about 4 million decimal digits
(exactly 4,053,946 digits).
Trying only to print this number 
$$(924947738006701322247758 \cdots 1130073855470256259071)$$
would need around 200 pages in the Financial Times.

@@

    The Fermat number
    $$ F[2,145,351] = 2^{(2^{2,145,351})} + 1 $$ 
    is divisible by the prime
    $$ p = 3*2^{2,145,353} + 1 $$ \\
    This prime p is the largest known prime generalized Mersenne 
    number\index{Mersenne!number!generalized}. 
    It was discovered only a few days after Michael Angel discovered the 
    largest generalized Fermat number (\hyperlink{generalizedFermatprimes}
    {see below}).  This moved GIMPS' \index{GIMPS} first prime discovery,
    M-35 = M1398269, into 7th place.

@@

For large $n$, $P(n)$  lies between  $2n$  and  $n^2$. This means that there are
fewer primes numbers than even natural numbers but more prime numbers than
square numbers.

@@

Under the heading {\em How Many Primes Are There} at
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://www.utm.edu/research/primes/howmany.shtml}
               {\tt http://www.utm.edu/research/primes/howmany.shtml}
\end{itemize}
you will find numerous other details.

@@

  \item The German \index{Riemann Bernhard} mathematician Bernhard Riemann
      (1826-1866) put forward a formula for the distribution of primes
      that would further improve the estimate. However, this has neither
      been proved nor contradicted so far.
\end{itemize}

@@

% --------------------------------------------------------------------------
\newpage
\subsubsection{Orders of magnitude / dimensions in reality}\label{s:grosord}
In the description of cryptographic protocols and algorithms, numbers occur that
are so large or so small that they are inaccessible to our intuitive
understanding. It may therefore be useful to provide comparative numbers from
the real world around us so that we can develop a feeling for the
security of cryptographic algorithms. Some of the numbers listed below originate
from \cite{Schwenk1996} and \cite[p.18]{Schneier1996p}.  \hypertarget{grosord}{}
\begin{tabbing}
Probability that you will be hijacked on your next flight:~~ \= abcdefhijk \=
\kill
Probability that you will be hijacked on your next flight \> $ 5.5 \cdot 10^{-6}
$\> \\
Probability of 6 correct numbers in the lottery \> $ 7.1 \cdot 10^{-8} $\> \\
Annual probability of being hit by lightning \> $ 10^{-7} $\> \\
Risk of being hit by a meteorite \> $ 1.6 \cdot 10^{-12} $ \> \\
--------------------------------------------------------------------------------
----------------------------------\\
Time until the next ice age (in years) \> $14000 $ \> $ (2^{14})$ \\
Time until the sun dies away (in years)   \> $10^{9} $ \> $(2^{30})$ \\
Age of the Earth (in years)\> $ 10^9 $ \> $(2^{30}) $  \\
Age of the universe (in years) \> $ 10^{10} $ \> $(2^{34}) $ \\
Number of the Earth's atoms            \> $10^{51} $ \> $ (2^{170}) $ \\
Number of the sun's atoms              \> $10^{57}$ \> $ (2^{190})$ \\
Number of atoms in the universe (without dark material)        \> $10^{77}$  \>
$ (2^{265})$ \\
Volume of the universe (in $cm^3$)     \> $10^{84}$ \> $(2^{280})$
\end{tabbing}

@@

\bibitem[Aaronson2003]{Aaronson2003} \index{Aaronson 2003}
    Scott Aaronson, \\
    {\em The Prime Facts: From Euclid to AKS}, \\
    \href{http://www.cs.berkeley.edu/~aaronson/prime.ps}
         {\texttt{http://www.cs.berkeley.edu/\~{}aaronson/prime.ps}}. \\
    Only after I had completed this article, I did come across the 
    extremely well-written paper by Scott Aaronson, which also offers
    a didactically well done intoduction to this topic. It is 
    humorous and easy to read but at the same time precise and erudite.

@@

I would like to take this opportunity to thank Mr. Henrik Koy and Mr. Roger
Oyono for their very constructive proof-reading of this article.

>>>>

This number was discovered as 39th Mersenne prime (and already called M-39,
despite it has not been proven yet, whether no further Mersenne prime
numbers between M-38 und M13466917 do indeed exist), $$2^{13,466,917}-1,$$
at December 6, 2001 -- more exactly, the verification of this number,
found at November 14, 2001 by the Canadian student Michael Cameron, was
successfully completed. This number has about 4 million decimal digits
(exactly 4,053,946 digits).
Trying only to print this number 
$$(924947738006701322247758 \cdots 1130073855470256259071)$$
would require around 200 pages in the Financial Times.

@@

    The Fermat number
    $$ F[2,145,351] = 2^{(2^{2,145,351})} + 1 $$ 
    is divisible by the prime
    $$ p = 3*2^{2,145,353} + 1 $$ \\
    This prime p is the largest known prime generalized Mersenne 
    number\index{Mersenne!number!generalized}. 
    It was discovered only a few days after Michael Angel has come across the 
    largest generalized Fermat number (\hyperlink{generalizedFermatprimes}
    {see below}).  This moved GIMPS' \index{GIMPS} first prime discovery,
    M-35 = M1398269, into 7th place.

@@

For large $n$, $P(n)$  lies between  $2n$  and  $n^2$. This means that there are
fewer prime numbers than even natural numbers but more prime numbers than
square numbers.

@@

Under the heading {\em How Many Primes Are There} at
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://www.utm.edu/research/primes/howmany.shtml}
               {\tt http://www.utm.edu/research/primes/howmany.shtml}
\end{itemize}
\vspace{-10pt}
you will find numerous other details.

@@

  \item The German \index{Riemann Bernhard} mathematician Bernhard Riemann
      (1826-1866) put forward a formula for the distribution of primes
      that would further improve the estimate. However, this has neither
      been proved nor disproved so far.
\end{itemize}

@@

% --------------------------------------------------------------------------
\newpage
\subsubsection{Orders of magnitude / dimensions in reality}\label{s:grosord}
In the description of cryptographic protocols and algorithms, numbers occur that
are so large or so small that they are inaccessible to our intuitive
understanding. It may therefore be useful to provide comparative numbers from
the real world around us so that we can develop a feeling for the
security of cryptographic algorithms. Some of the numbers listed below originate
from \cite{Schwenk1996} and \cite[p.18]{Schneier1996p}.  \hypertarget{grosord}{}
\begin{tabbing}
Probability that you will be hijacked on your next flight:~~ \= abcdefhijk \=
\kill
Probability that you will be hijacked on your next flight \> $ 5.5 \cdot 10^{-6}
$\> \\
Probability of 6 correct numbers in the lottery \> $ 7.1 \cdot 10^{-8} $\> \\
Annual probability of being hit by lightning \> $ 10^{-7} $\> \\
Risk of being hit by a meteorite \> $ 1.6 \cdot 10^{-12} $ \> \\
--------------------------------------------------------------------------------
----------------------------------\\
Time until the next ice age (in years) \> $14000 $ \> $ (2^{14})$ \\
Time until the sun dies (in years)   \> $10^{9} $ \> $(2^{30})$ \\
Age of the Earth (in years)\> $ 10^9 $ \> $(2^{30}) $  \\
Age of the universe (in years) \> $ 10^{10} $ \> $(2^{34}) $ \\
Number of the Earth's atoms            \> $10^{51} $ \> $ (2^{170}) $ \\
Number of the sun's atoms              \> $10^{57}$ \> $ (2^{190})$ \\
Number of atoms in the universe (without dark material)        \> $10^{77}$  \>
$ (2^{265})$ \\
Volume of the universe (in $cm^3$)     \> $10^{84}$ \> $(2^{280})$
\end{tabbing}

@@

\bibitem[Aaronson2003]{Aaronson2003} \index{Aaronson 2003}
    Scott Aaronson, \\
    {\em The Prime Facts: From Euclid to AKS}, \\
    \href{http://www.cs.berkeley.edu/~aaronson/prime.ps}
         {\texttt{http://www.cs.berkeley.edu/\~{}aaronson/prime.ps}}. \\
    Only after I had completed this article, did I come across the 
    extremely well-written paper by Scott Aaronson, which also offers
    a didactically well done intoduction to this topic. It is 
    humorous and easy to read but at the same time precise and erudite.

@@

I would like to take this opportunity to thank Mr.\ Henrik Koy and Mr.\ Roger
Oyono for their very constructive proof-reading of this article.
